---
title: "Risk Model for Human Challenge Trials"
author: "1DaySooner"
output:
  flexdashboard::flex_dashboard:
    orientation: rows
    source_code: embed
runtime: shiny
resource_files:
- Riskmodel_Beta.html
---

```{r global, include=FALSE}
library(ggplot2)
library(shiny) # I see that it says runtime:Shiny, but without this line I get errors.
library(scales)
library(tidyverse)
library(flexdashboard)
source("Riskmodel_Beta.R")


# Function Definitions (from Shiny_Alpha.R)
mm_tribble <- tribble(
      ~activity, ~micromorts, ~col,
      "Skydiving Once", 8, "Other",
      "Running a Marathon", 7, "Other",
      "Motorcycle from NYC to DC", 37.5, "Other",
      #"Motorcycle Trip \n from NYC to LA", 465, "Other",
      "Living Kidney Donation",300, "Other",
      "Giving birth", 60, "Other",
      #"Soldier in Afganistan \n in 2009 for 1 week", 336, "Other",
      "Trucking for 1 year", 280, "Other",
      #"Logging for 1 year", 737, "Other",
      "Liver Donation \n (left lobe)", 1000, "Other")

micromortify <- function(Micromorts){
  mm_names <- unlist(mm_tribble['activity'])
  mm_values <- unlist(mm_tribble['micromorts'])
  closest_answer <- (abs(mm_values - Micromorts) == min(abs(mm_values - Micromorts)))
  output_list <- list(name = mm_names[closest_answer], num = mm_values[closest_answer])
}

basicFormatter <- function(Number, Mode='Pct', display_digits=3) {
  if (Mode == 'Pct') {
    return(
      paste0(
        prettyNum(Number * 100, digits = display_digits, format = "g", drop0trailing = TRUE), "%")
    )
  }
}
```

Inputs {.sidebar}
=======================================================================

```{r}
# Make the sliders

sliderInput(
  inputId = 'num_participants', 
  label = 'Number of Trial Participants', 
  min = 1, 
  max = 100,
  value = 50,
  step = 1
)

sliderInput(
  inputId = 'p_treatment', 
  label = 'Potential Treatment Efficacy', 
  min = 0, 
  max = 0.9,
  value = 0,
  step = 0.05
)

selectInput(
  'gender', 
  'Gender of Participants', 
  choices = c("Male" = "m", "Female" = "f", "Both (Equal Number)" = "b"),
  selected = "b"
)

selectInput(
  'ages', 
  'Ages of Participants', 
  c("20 to 29", "20 to 39"), 
  selected = "20 to 29"
)

checkboxInput(
'adjustment_status', 
'Adjust for no comorbidities', 
value = TRUE, 
width = NULL
)

checkboxInput(
'adjustment_population', 
'Adjust for best-case population', 
value = TRUE, 
width = NULL
)


```


<!-- Notice the ====== vs. -------. These just change how the page is divided up into sections vs. rows -->

Dashboard
=======================================================================

Row
-----------------------------------------------------------------------

### Activity with Comparable Risk to Individual Participation in Study

```{r}

mm_ify = reactive(micromortify(micromorts()))

renderValueBox({
  mm_name <- mm_ify()$name
  valueBox(
    value = paste0("Risk similar to: ", mm_name),
    color = c("#8e8db0")
  )
})
```

Row
-----------------------------------------------------------------------

### Probability of no deaths during study {.value-box}

```{r}

# defines a reactive "function" called outcome_mean. When you want the up to date value, just call outcome_mean() and it'll give you whatever Simulate_StudyRisks returns. It knows where Simulate_StudyRisks is because of the  source("Riskmodel_Alpha.R") line we put at the top of the page.
deaths_mean <-
  reactive(
    1-StudyRisk(
      Participants = input$num_participants,
      Age_Range = input$ages,
      gender = input$gender,
      Therapy = input$p_treatment,
            sta = (if (input$adjustment_status) "healthy" else "all"),
      population = (if (input$adjustment_population) "best" else "meta"),
      Pctile = '50%',
      outcome = "death"
    )
  )

deaths_low <-
  reactive(
    1 - StudyRisk(
      Participants = input$num_participants,
      Age_Range = input$ages,
      gender = input$gender,
      Therapy =  1-((1-input$p_treatment)),
      sta = (if (input$adjustment_status) "healthy" else "all"),
      population = (if (input$adjustment_population) "best" else "meta"),
      Pctile="2.5%",
      outcome = "death")
  )

deaths_high <-
  reactive(
    1 - StudyRisk(
      Participants = input$num_participants,
      Age_Range = input$ages,
      gender = input$gender,
      Therapy =  1-((1-input$p_treatment)),
      sta = (if (input$adjustment_status) "healthy" else "all"),
      population = (if (input$adjustment_population) "best" else "meta"),
      Pctile="97.5%",
      outcome = "death")
  )

hosp_mean <-
  reactive(
    1-StudyRisk(
      Participants = input$num_participants,
      Age_Range = input$ages,
      gender = input$gender,
      Therapy =  1-((1-input$p_treatment)),
            sta = (if (input$adjustment_status) "healthy" else "all"),
      population = (if (input$adjustment_population) "best" else "meta"),
      Pctile = '50%',
      outcome = "hosp"
    )
  )
hosp_low <-
  reactive(
    1-StudyRisk(
      Participants = input$num_participants,
      Age_Range = input$ages,
      gender = input$gender,
      Therapy =  1-((1-input$p_treatment)),
            sta = (if (input$adjustment_status) "healthy" else "all"),
      population = (if (input$adjustment_population) "best" else "meta"),
      Pctile = '2.5%',
      outcome = "hosp"
    )
  )

hosp_high <-
  reactive(
    1-StudyRisk(
      Participants = input$num_participants,
      Age_Range = input$ages,
      gender = input$gender,
      Therapy =  1-((1-input$p_treatment)),
            sta = (if (input$adjustment_status) "healthy" else "all"),
      population = (if (input$adjustment_population) "best" else "meta"),
      Pctile = '97.5%',
      outcome = "hosp"
    )
)



# this is where we call the actual reactive "function". There are different kinds of renders, such as renderValueBox, renderPlot, etc. Just Google them to see how to use them/what they look like

#bonus, you might consider adding icons, or having color change dynamically as the value changes (e.g. red for high risk, green for low risk). This can also be easily googled.
renderValueBox({
  output = deaths_mean()
  out_mean <- basicFormatter(output, display_digits=3)
  if (output > 0.99945) out_mean <- basicFormatter(output, display_digits=5) else if (output > 0.9945) out_mean <- basicFormatter(output, display_digits=4)
  valueBox(
    value = out_mean,
    caption = paste0("Probability of no deaths during study. \n 95% CI: ", basicFormatter(deaths_high(), display_digits=5), " to ", basicFormatter(deaths_low(), display_digits=5)),
    # icon = "icon_name"
    color = c("#45497d")
  )
})
```

### Probability of no hospitalizations during study {.value-box}
```{r}

renderValueBox({
  output = hosp_mean()
  out_mean <- basicFormatter(output, display_digits=3)
  if (output > 0.9945) out_mean <- basicFormatter(output, display_digits=4)
  valueBox(
    value = out_mean,
        caption = paste0("Probability of no deaths during study. \n 95% CI: ", basicFormatter(hosp_high(), display_digits=4), " to ", basicFormatter(hosp_low(), display_digits=4)),
    # icon = "icon_name"
    color = c("#45497d")
  )
})
```


```{r}
deaths_95 <-
  reactive(
    1 - StudyRisk(
      Participants = input$num_participants,
      Age_Range = input$ages,
      gender = input$gender,
      Therapy =  1-((1-input$p_treatment)),
      sta = (if (input$adjustment_status) "healthy" else "all"),
      population = (if (input$adjustment_population) "best" else "meta"),
      Pctile="95%",
      outcome = "death")
  )

hosp_95 <- #Needed later.
  reactive(
    1 - StudyRisk(
      Participants = input$num_participants,
      Age_Range = input$ages,
      gender = input$gender,
      Therapy =  1-((1-input$p_treatment)),
      sta = (if (input$adjustment_status) "healthy" else "all"),
      population = (if (input$adjustment_population) "best" else "meta"),
      Pctile="95%",
      outcome = "hosp")
  )

```

### Micromorts {.value-box}

```{r}
micromorts <-
  reactive(
    round(
      IndivRiskPull(
        Age_Range = input$ages,
        gender = input$gender,
        outcome = 'death',
        Therapy = input$p_treatment, 
        sta = (if (input$adjustment_status) "healthy" else "all"),
        population = (if (input$adjustment_population) "best" else "meta"),
        Pctile = "50%"
      ) * 1000000
    )
  )

micromorts_low <-
  reactive(
    round(
      IndivRiskPull(
        Age_Range = input$ages,
        gender = input$gender,
        outcome = 'death',
        Therapy = input$p_treatment, 
        sta = (if (input$adjustment_status) "healthy" else "all"),
        population = (if (input$adjustment_population) "best" else "meta"),
        Pctile = "2.5%"
      ) * 1000000,1
    )
  )

micromorts_high <-
  reactive(
    round(
      IndivRiskPull(
        Age_Range = input$ages,
        gender = input$gender,
        outcome = 'death',
        Therapy = input$p_treatment, 
        sta = (if (input$adjustment_status) "healthy" else "all"),
        population = (if (input$adjustment_population) "best" else "meta"),
        Pctile = "97.5%"
      ) * 1000000,1
    )
  )


renderValueBox({
  mm <- micromorts()
  valueBox(
    value = mm,
    caption = paste0("Micromorts (unit of risk, one-in-a-million chance of death) per individual. \n 95% CI: ", micromorts_low(), " to ", micromorts_high()),
    color = c("#45497d")
  )
})
```

Row {.tabset}
-----------------------------------------------------------------------

### % Risk

```{r}
graph_risks <-
  reactive(c(IndivRiskPull(
        Age_Range = input$ages,
        gender = input$gender,
        outcome = 'death',
        Therapy =  1-((1-input$p_treatment)), 
        sta = (if (input$adjustment_status) "healthy" else "all"),
        population = (if (input$adjustment_population) "best" else "meta"),
        Pctile = "50%"),
      IndivRiskPull(
        Age_Range = input$ages,
        gender = input$gender,
        outcome = 'hosp',
        Therapy =  1-((1-input$p_treatment)), 
        sta = (if (input$adjustment_status) "healthy" else "all"),
        population = (if (input$adjustment_population) "best" else "meta"),
        Pctile = "50%"),
      IndivRiskPull(
        Age_Range = input$ages,
        gender = input$gender,
        outcome = 'death',
        Therapy =  1-((1-input$p_treatment)), 
        sta = (if (input$adjustment_status) "healthy" else "all"),
        population = (if (input$adjustment_population) "best" else "meta"),
        Pctile = "95%"),
      IndivRiskPull(
        Age_Range = input$ages,
        gender = input$gender,
        outcome = 'hosp',
        Therapy =  1-((1-input$p_treatment)),
        sta = (if (input$adjustment_status) "healthy" else "all"),
        population = (if (input$adjustment_population) "best" else "meta"),
        Pctile = "95%")
  )
  )

renderPlot({
    out_expected <- 
    tribble(
      ~out_mean, ~status,
      graph_risks()[1], "Death",
      graph_risks()[2], "Hospitalized",
    )
    
  out_95_percent <- 
    tribble(
      ~out_95, ~status,
      graph_risks()[3], "Death",
      graph_risks()[4], "Hospitalized",
    )
  
  # the code below might look weird but it's just vanilla ggplot2. https://dcl-data-vis.stanford.edu/ggplot2-basics.html is a decent start, https://r4ds.had.co.nz/data-visualisation.html is also a great resource!!
  out_95_percent %>% 
    ggplot(aes(status, out_95)) +
    geom_col(fill =c("#1B2B65")) +
    scale_y_continuous(labels = scales::label_percent(scale = 10), breaks = scales::breaks_pretty(7), limits = c(0, 0.01)) +
    labs(
      title = "Risk by Outcome",
      subtitle = "95% probability that actual value is in contained within the blue shaded region",
      x = NULL,
      y = "% Risk",
      color = "Expected"
    ) +
    theme_minimal() 
})
```

### Micromorts

```{r}
m_micromorts <-
  reactive(
    round(
      IndivRiskPull(
        Age_Range = input$ages,
        gender = 'm',
        outcome = 'death',
        Therapy =  1-((1-input$p_treatment)), 
        sta = (if (input$adjustment_status) "healthy" else "all"),
        population = (if (input$adjustment_population) "best" else "meta"),
        Pctile = "50%"
      ) * 1000000
    )
  )

f_micromorts <-
  reactive(
    round(
      IndivRiskPull(
        Age_Range = input$ages,
        gender = 'f',
        outcome = 'death',
        Therapy =  1-((1-input$p_treatment)),
        sta = (if (input$adjustment_status) "healthy" else "all"),
        population = (if (input$adjustment_population) "best" else "meta"),
        Pctile = "50%"
      ) * 1000000
    )
  )

b_micromorts <-
  reactive(
    round(
      IndivRiskPull(
        Age_Range = input$ages,
        gender = 'b',
        outcome = 'death',
        Therapy =  1-((1-input$p_treatment)),
        sta = (if (input$adjustment_status) "healthy" else "all"),
        population = (if (input$adjustment_population) "best" else "meta"),
        Pctile = "50%"
      ) * 1000000
    )
  )

renderPlot({
  mm_tribble %>% 
    add_row(activity="HCT Volunteer (Male)", micromorts=m_micromorts(), col="Participate") %>%
    add_row(activity="HCT Volunteer (Female)", micromorts=f_micromorts(), col="Participate") %>%
    ggplot(aes(fct_reorder(activity, micromorts, .desc = FALSE), micromorts, fill=col)) +
    geom_col() +
    labs(
      title = "Micromorts by Gender",
      x = NULL,
      y = "Micromorts"
    ) +
    theme_minimal() +
    scale_fill_manual(values=c("#1B2B65", "#FFD600")) +
    theme(legend.position = "none")
})
```

About Model
=======================================================================

<h3>Using the Model</h3>
This model represents the risk of deaths or hospitalizations during a notional Human Challenge Trial dosing study for COVID-19, of the type <a href="https://edition.cnn.com/2020/10/20/health/uk-coronavirus-challenge-study-intl/index.html">planned in the UK in early 2021</a>. It displays the probability that there will be no deaths and no hospitalizations for all participants in the study, along with the number of "micromorts" which are faced by each partipant in the study. Below this, a roughly comparable risk of a different type is shown.<br />

The interface allows choosing whether to adjust for expected-case mortality, based on looking at a population with no comorbidities, using estimates from the location in the meta-analysis with the best possible care and treatment. It also allows varying the characteristics and number of participants to view overall risks of the study.<br />

Note that this model is appropriate for considering a dosage trial, and while the size of such a trial may vary, one approach would be to expose 5 volunteers each to 3 increasing titers of COVID-19, stopping if the majority or all volunteers for a given exposure level develop clinical disease. Following this, an additional 10-20 volunteers would be given the appropriate titer of COVID-19, to validate the initial result.  The default number of participants is therefor 50.<br />

<h3>Methodology</h3>
  The methodology for this preliminary, not peer reviewed risk model is detailed in a forthcoming paper. <a href='https://github.com/1DaySooner/RiskModel/blob/master/ifr-model/bayesian_ifr_model.pdf'>The appendix of the working paper detailing the model</a> can be found in <a href="https://github.com/1DaySooner/RiskModel/">our github repository</a>, along with source code and other details. This model estimates IFR, and hospitalization rates are based on relative risk of hospitalization and mortality from <a hre="https:// science.sciencemag.org/cgi/content/full/science.abc3517">Salje et al.</a>
  
<!-- This doesn't work. "Not Found". Grrr.
Preprint Paper
=======================================================================

```{r}
renderUI({
  tags$iframe(style="height:600px; width:100%", src="./bayesian_ifr_model.pdf") 
  })
```
-->