---
title: "Risk Model for Human Challenge Trials"
author: "by 1DaySooner"
output:
  flexdashboard::flex_dashboard:
    orientation: rows
    source_code: embed
runtime: shiny
resource_files:
- Riskmodel.html
---

```{r global, include=FALSE}
library(ggplot2)
library(shiny) # I see that it says runtime:Shiny, but without this line I get errors.
library(scales)
library(kableExtra)
library(tidyverse)
library(flexdashboard)
source("Riskmodel.R")




# Function Definitions (from Shiny_Alpha.R)
mm_tribble <- tribble(
      ~activity, ~micromorts, ~col,
      "Running a Marathon", 7, "Other",
      # "Skydiving Once", 8, "Other",
      # "Motorcycle trip from NYC to DC", 37.5, "Other",
      #"Motorcycle Trip \n from NYC to LA", 465, "Other",
      "Giving birth (UK)", 60, "Other",
      "Liposuction", 200, "Other",
      "Living Kidney Donation",300, "Other",
      #"Soldier in Afganistan \n in 2009 for 1 week", 336, "Other",
      #"Trucking for 1 year", 280, "Other",
      #"Logging for 1 year", 737, "Other",
      "Liver Donation \n (left lobe)", 1000, "Other"
      )

micromortify <- function(Micromorts){
  mm_names <- unlist(mm_tribble['activity'])
  mm_values <- unlist(mm_tribble['micromorts'])
  next_larger <- which(mm_values - Micromorts > 0)[1]
  #closest_answer <- (abs(mm_values - Micromorts) == min(abs(mm_values - Micromorts)))
  output_list <- list(name = mm_names[next_larger], num = mm_values[next_larger], rr = Micromorts / mm_values[next_larger])
}

basicFormatter <- function(Number, Mode='Pct', display_digits=3) {
  if (Mode == 'Pct') {
    return(
      paste0(
        prettyNum(Number * 100, digits = display_digits, format = "g", drop0trailing = TRUE), "%")
    )
  }
  if (Mode == 'Odds') {
    in_X = round(-log10(Number)+1)
    Y_in = round(Number*(10^in_X),1)
    return(
      paste0(Y_in, " in ", format(10^in_X, big.mark=",", scientific=FALSE))
    )
  }
}


```

Inputs {.sidebar}
=======================================================================

<b>Trial characteristics:</b>

```{r}
# Make the sliders

sliderInput(
  inputId = 'num_participants', 
  label = 'Number of Trial Participants', 
  min = 1, 
  max = 100,
  value = 50,
  step = 1
)

selectInput(
  'gender', 
  'Gender of Participants', 
  choices = c("Male" = "m", "Female" = "f", "Both (Equal Number)" = "b"),
  selected = "b"
)

selectInput(
  'ages', 
  'Ages of Participants', 
  c("20 to 29", "20 to 39"), 
  selected = "20 to 29"
)


```

<b>Risk adjustment for:</b>

```{r}

sliderInput(
  inputId = 'p_treatment', 
  label = 'Potential Treatment Efficacy', 
  min = 0, 
  max = 0.9,
  value = 0,
  step = 0.05
)

checkboxInput(
'adjustment_status', 
'No comorbidities', 
value = TRUE, 
width = NULL
)

checkboxInput(
'adjustment_population', 
'Population heterogenity', 
value = FALSE, 
width = NULL
)

```

The adjustment for potential treatment, comorbidities, and population heterogeneity are explained in the "About Model" tab.

<b>This does not include an estimate of longer term risks.</b> See the discussion in the "Long Term Risks" tab for more information.

<!-- Notice the ====== vs. -------. These just change how the page is divided up into sections vs. rows -->

Dashboard
=======================================================================

Row
-----------------------------------------------------------------------

### Comparable Risk to Individual Participation in Study

```{r}

mm_ify = reactive(micromortify(micromorts()))

renderValueBox({
  mm_name <- mm_ify()$name
  valueBox(
    value = paste0("Participation is ", basicFormatter(mm_ify()$rr,"Pct",1), " as risky as ", mm_ify()$name),
    color = c("#8e8db0")
    )
})
```
<!--- Doesn't work - displays NAN.
### Risk Gauge

```{r}
renderGauge({
  gauge(value=(mm_ify()$rr), min = 0, max = 1, symbol = '%')
})
```
--->

Row
-----------------------------------------------------------------------

### Average probability of no deaths during study {.value-box}

```{r}

# defines a reactive "function" called outcome_mean. When you want the up to date value, just call outcome_mean() and it'll give you whatever Simulate_StudyRisks returns. It knows where Simulate_StudyRisks is because of the  source("Riskmodel_Alpha.R") line we put at the top of the page.
deaths_mean <-
  reactive(
    1-StudyRisk(
      Participants = input$num_participants,
      Age_Range = input$ages,
      gender = input$gender,
      Therapy = input$p_treatment,
            sta = (if (input$adjustment_status) "healthy" else "all"),
      population = (if (input$adjustment_population) "best" else "meta"),
      Pctile = '50%',
      outcome = "death"
    )
  )

deaths_low <-
  reactive(
    1 - StudyRisk(
      Participants = input$num_participants,
      Age_Range = input$ages,
      gender = input$gender,
      Therapy =  1-((1-input$p_treatment)),
      sta = (if (input$adjustment_status) "healthy" else "all"),
      population = (if (input$adjustment_population) "best" else "meta"),
      Pctile="2.5%",
      outcome = "death")
  )

deaths_high <-
  reactive(
    1 - StudyRisk(
      Participants = input$num_participants,
      Age_Range = input$ages,
      gender = input$gender,
      Therapy =  1-((1-input$p_treatment)),
      sta = (if (input$adjustment_status) "healthy" else "all"),
      population = (if (input$adjustment_population) "best" else "meta"),
      Pctile="97.5%",
      outcome = "death")
  )

hosp_mean <-
  reactive(
    1-StudyRisk(
      Participants = input$num_participants,
      Age_Range = input$ages,
      gender = input$gender,
      Therapy =  1-((1-input$p_treatment)),
            sta = (if (input$adjustment_status) "healthy" else "all"),
      population = (if (input$adjustment_population) "best" else "meta"),
      Pctile = '50%',
      outcome = "hosp"
    )
  )
hosp_low <-
  reactive(
    1-StudyRisk(
      Participants = input$num_participants,
      Age_Range = input$ages,
      gender = input$gender,
      Therapy =  1-((1-input$p_treatment)),
            sta = (if (input$adjustment_status) "healthy" else "all"),
      population = (if (input$adjustment_population) "best" else "meta"),
      Pctile = '2.5%',
      outcome = "hosp"
    )
  )

hosp_high <-
  reactive(
    1-StudyRisk(
      Participants = input$num_participants,
      Age_Range = input$ages,
      gender = input$gender,
      Therapy =  1-((1-input$p_treatment)),
            sta = (if (input$adjustment_status) "healthy" else "all"),
      population = (if (input$adjustment_population) "best" else "meta"),
      Pctile = '97.5%',
      outcome = "hosp"
    )
)



# this is where we call the actual reactive "function". There are different kinds of renders, such as renderValueBox, renderPlot, etc. Just Google them to see how to use them/what they look like

#bonus, you might consider adding icons, or having color change dynamically as the value changes (e.g. red for high risk, green for low risk). This can also be easily googled.
renderValueBox({
  output = deaths_mean()
  out_mean <- basicFormatter(output, display_digits=3)
  if (output > 0.99945) out_mean <- basicFormatter(output, display_digits=5) else if (output > 0.9945) out_mean <- basicFormatter(output, display_digits=4)
  valueBox(
    value = out_mean,
    caption = paste0("Average probability of no deaths during study. \n 95% CI: ", basicFormatter(deaths_high(), display_digits=5), " to ", basicFormatter(deaths_low(), display_digits=5)),
    # icon = "icon_name"
    color = c("#45497d")
  )
})
```

### Average probability of no hospitalizations during study {.value-box}
```{r}

renderValueBox({
  output = hosp_mean()
  out_mean <- basicFormatter(output, display_digits=3)
  if (output > 0.9945) out_mean <- basicFormatter(output, display_digits=4)
  valueBox(
    value = out_mean,
        caption = paste0("Average probability of no cases requiring hospitalization during study. \n 95% CI: ", basicFormatter(hosp_high(), display_digits=4), " to ", basicFormatter(hosp_low(), display_digits=4)),
    # icon = "icon_name"
    color = c("#45497d")
  )
})
```


```{r}
deaths_95 <-
  reactive(
    1 - StudyRisk(
      Participants = input$num_participants,
      Age_Range = input$ages,
      gender = input$gender,
      Therapy =  1-((1-input$p_treatment)),
      sta = (if (input$adjustment_status) "healthy" else "all"),
      population = (if (input$adjustment_population) "best" else "meta"),
      Pctile="95%",
      outcome = "death")
  )

hosp_95 <- #Needed later.
  reactive(
    1 - StudyRisk(
      Participants = input$num_participants,
      Age_Range = input$ages,
      gender = input$gender,
      Therapy =  1-((1-input$p_treatment)),
      sta = (if (input$adjustment_status) "healthy" else "all"),
      population = (if (input$adjustment_population) "best" else "meta"),
      Pctile="95%",
      outcome = "hosp")
  )

```

### Average probability of Individual Death {.value-box}

```{r}
Odds <-
  reactive(
    basicFormatter(
      IndivRiskPull(
        Age_Range = input$ages,
        gender = input$gender,
        outcome = 'death',
        Therapy = input$p_treatment, 
        sta = (if (input$adjustment_status) "healthy" else "all"),
        population = (if (input$adjustment_population) "best" else "meta"),
        Pctile = "50%"
      ), "Odds"
    )
  )

Odds_low <-
  reactive(
    basicFormatter(
      IndivRiskPull(
        Age_Range = input$ages,
        gender = input$gender,
        outcome = 'death',
        Therapy = input$p_treatment, 
        sta = (if (input$adjustment_status) "healthy" else "all"),
        population = (if (input$adjustment_population) "best" else "meta"),
        Pctile = "2.5%"
      ), "Odds"
    )
  )


Odds_high <-
  reactive(
    basicFormatter(
      IndivRiskPull(
        Age_Range = input$ages,
        gender = input$gender,
        outcome = 'death',
        Therapy = input$p_treatment, 
        sta = (if (input$adjustment_status) "healthy" else "all"),
        population = (if (input$adjustment_population) "best" else "meta"),
        Pctile = "97.5%"
      ), "Odds"
    )
  )


micromorts <-
  reactive(
    round(
      IndivRiskPull(
        Age_Range = input$ages,
        gender = input$gender,
        outcome = 'death',
        Therapy = input$p_treatment, 
        sta = (if (input$adjustment_status) "healthy" else "all"),
        population = (if (input$adjustment_population) "best" else "meta"),
        Pctile = "50%"
      ) * 1000000,1
    )
  )


micromorts_low <-
  reactive(
    round(
      IndivRiskPull(
        Age_Range = input$ages,
        gender = input$gender,
        outcome = 'death',
        Therapy = input$p_treatment, 
        sta = (if (input$adjustment_status) "healthy" else "all"),
        population = (if (input$adjustment_population) "best" else "meta"),
        Pctile = "2.5%"
      ) * 1000000,1
    )
  )

micromorts_high <-
  reactive(
    round(
      IndivRiskPull(
        Age_Range = input$ages,
        gender = input$gender,
        outcome = 'death',
        Therapy = input$p_treatment, 
        sta = (if (input$adjustment_status) "healthy" else "all"),
        population = (if (input$adjustment_population) "best" else "meta"),
        Pctile = "97.5%"
      ) * 1000000,1
    )
  )


renderValueBox({
  mm <- micromorts()
  valueBox(
    value = Odds(),
    caption = paste0("Average odds of Individual Death. 95% CI: ", Odds_low(), " to ", Odds_high()),
    color = c("#45497d")
  )
})
```

Row {.tabset}
-----------------------------------------------------------------------

### % Risk

```{r}
graph_risks <-
  reactive(c(IndivRiskPull(
        Age_Range = input$ages,
        gender = input$gender,
        outcome = 'death',
        Therapy =  1-((1-input$p_treatment)), 
        sta = (if (input$adjustment_status) "healthy" else "all"),
        population = (if (input$adjustment_population) "best" else "meta"),
        Pctile = "50%"),
      IndivRiskPull(
        Age_Range = input$ages,
        gender = input$gender,
        outcome = 'hosp',
        Therapy =  1-((1-input$p_treatment)), 
        sta = (if (input$adjustment_status) "healthy" else "all"),
        population = (if (input$adjustment_population) "best" else "meta"),
        Pctile = "50%"),
      IndivRiskPull(
        Age_Range = input$ages,
        gender = input$gender,
        outcome = 'death',
        Therapy =  1-((1-input$p_treatment)), 
        sta = (if (input$adjustment_status) "healthy" else "all"),
        population = (if (input$adjustment_population) "best" else "meta"),
        Pctile = "95%"),
      IndivRiskPull(
        Age_Range = input$ages,
        gender = input$gender,
        outcome = 'hosp',
        Therapy =  1-((1-input$p_treatment)),
        sta = (if (input$adjustment_status) "healthy" else "all"),
        population = (if (input$adjustment_population) "best" else "meta"),
        Pctile = "95%")
  )
  )

renderPlot({
    out_expected <- 
    tribble(
      ~out_mean, ~status,
      graph_risks()[1], "Death",
      graph_risks()[2], "Hospitalized",
    )
    
  out_95_percent <- 
    tribble(
      ~out_95, ~status,
      graph_risks()[3], "Death",
      graph_risks()[4], "Hospitalized",
    )
  
  # the code below might look weird but it's just vanilla ggplot2. https://dcl-data-vis.stanford.edu/ggplot2-basics.html is a decent start, https://r4ds.had.co.nz/data-visualisation.html is also a great resource!!
  out_95_percent %>% 
    ggplot(aes(status, out_95)) +
    geom_col(fill =c("#1B2B65")) +
    scale_y_continuous(labels = scales::label_percent(scale = 10), breaks = scales::breaks_pretty(7), limits = c(0, 10^(round(log10(graph_risks()[4]))+0.5))) +
    labs(
      title = "Risk by Outcome",
      subtitle = "95% probability that actual value is in contained within the blue shaded region",
      x = NULL,
      y = "% Risk",
      color = "Expected"
    ) + theme_minimal(base_size = 18) + theme(panel.grid.major.x = element_blank(), panel.grid.minor.x = element_blank())
})
```


### Micromorts

Micromorts are a standard measure of mortality risk defined as a 1-in-1 million chance of death.

```{r}
m_micromorts <-
  reactive(
    round(
      IndivRiskPull(
        Age_Range = input$ages,
        gender = 'm',
        outcome = 'death',
        Therapy =  1-((1-input$p_treatment)), 
        sta = (if (input$adjustment_status) "healthy" else "all"),
        population = (if (input$adjustment_population) "best" else "meta"),
        Pctile = "50%"
      ) * 1000000
    )
  )

f_micromorts <-
  reactive(
    round(
      IndivRiskPull(
        Age_Range = input$ages,
        gender = 'f',
        outcome = 'death',
        Therapy =  1-((1-input$p_treatment)),
        sta = (if (input$adjustment_status) "healthy" else "all"),
        population = (if (input$adjustment_population) "best" else "meta"),
        Pctile = "50%"
      ) * 1000000
    )
  )

b_micromorts <-
  reactive(
    round(
      IndivRiskPull(
        Age_Range = input$ages,
        gender = 'b',
        outcome = 'death',
        Therapy =  1-((1-input$p_treatment)),
        sta = (if (input$adjustment_status) "healthy" else "all"),
        population = (if (input$adjustment_population) "best" else "meta"),
        Pctile = "50%"
      ) * 1000000
    )
  )


max_y <- 500

renderPlot({
  mm_tribble %>% 
  add_row(activity="Challenge Volunteer (Male)", micromorts=m_micromorts(), col="Participate") %>%  
  add_row(activity="Challenge Volunteer (Female)", micromorts=f_micromorts(), col="Participate") %>% 
  mutate(label_y = ifelse(micromorts > max_y, max_y, micromorts + 25)) %>%
  ggplot(aes(fct_reorder(activity, micromorts, .desc = FALSE), micromorts, fill=col)) +
  geom_col() +
  labs(
    title = "Micromorts by Gender",
    x = NULL,
    y = "Micromorts"
  ) +
  theme_minimal(base_size = 18) + theme(panel.grid.major.x = element_blank(), panel.grid.minor.x = element_blank()) +
  scale_fill_manual(values=c("#1B2B65", "#FFD600")) +
  coord_cartesian(ylim = c(0,max_y)) + 
  geom_label(aes(label = micromorts, y = label_y), col = "white") +
  theme(legend.position = "none", axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1))
})
```

About Model
=======================================================================

<h3>Using the Model</h3>
This model represents the risk of deaths or hospitalizations during a notional Human Challenge Trial dosing study for COVID-19, of the type <a href="https://edition.cnn.com/2020/10/20/health/uk-coronavirus-challenge-study-intl/index.html">planned in the UK in early 2021</a>. It displays the average probability that there will be no deaths and no hospitalizations for all participants in the study, along with the number of "micromorts" which are faced by each participant in the study. Below this, a roughly comparable risk of a different type is shown.<br />

The interface allows choosing whether to adjust for expected-case mortality, based on looking at a population with no comorbidities, using estimates from the location in the meta-analysis with the best possible care and treatment. It also allows varying the characteristics and number of participants to view overall risks of the study.<br />

Note that this model is appropriate for considering a dosage trial, and while the size of such a trial may vary, one approach would be to expose 5 volunteers each to 3 increasing titers of COVID-19, stopping if the majority or all volunteers for a given exposure level develop clinical disease. Following this, an additional 10-20 volunteers would be given the appropriate titer of COVID-19, to validate the initial result.  The default number of participants is therefor 50.<br />

<h3>Methodology</h3>
The underlying model estimates IFR, and hospitalization rates are based on relative risk of hospitalization and mortality from <a href="https:// science.sciencemag.org/cgi/content/full/science.abc3517">Salje et al.</a> The methodology for this preliminary, not peer reviewed risk model is detailed in a forthcoming paper, available in the "PReprint" tab above. Further details of the model, and source code for the app, can be found in <a href="https://github.com/1DaySooner/RiskModel/">our github repository</a>, along with other details.

<h4>Thank you</h4>
The 1Day Sooner Research team thanks Nir Eyal and Sophie Rose for feedback on this model, and Emily Xu, Alex Hu, and the rest of the Develop for Good team for technical and graphic design assistance.
  

Long Term Risks
=======================================================================
<div class="LTR" width='100%' height='800'>
<iframe src='https://1daysooner.org/long-term-risks-faq#page' width='100%' height='600px'></iframe>
</div>


Preprint
=======================================================================
<div class="LTR" width='100%' height='800'>
<iframe src='./Exploring_Risks_of_Human_Challenge_Trials_for_COVID_19.pdf' width='100%' height='800'></iframe>
</div>

<!-- Include once we have a better format for citing them.
Data Sources for IFR Model
=======================================================================
```{r input-data-big}

df_tab <- read.csv("source_data.csv", row.names=1)

kbl(df_tab,
      col.names = c("Study location", "Scale", "Type", "End date", "Age", 
                    "Deaths", "N", "Mean", "2.5%", "97.5%", "IFR/100k"))

```
-->

